---
title: "Lab 5"
author: "David Wilson • dwils152@unc.edu"
date: "3/17/2022"
output: html_document
---

#### 1) This question uses data from this paper: https://science.sciencemag.org/content/347/6217/78. 
##### ***Variation in cancer risk among tissues can be explained by the number of stem cell divisions. Science  02 Jan 2015: Vol. 347, Issue 6217, pp. 78-81***

***

##### a) Download the data from here examining the relationship between the number of cell divisions and cancer risk: https://fodorclasses.github.io/classes/stats2020/cancerRisk.txt On a log10-log10 scale graph Lifetime_cancer_risk (on the y-axis) vs. CumulativeCellDivisions (on the x-axis).  This reproduces Fig. 1 from the paper. You can read in the file with read.table("cancerRisk.txt", header=TRUE, sep="\t")

```{r}

#cancer_risk <- read.table('cancerRisk.txt')

```

##### b) Using the lm function, fit a linear model with Lifetime_cancer_risk as the Y variable and CumulativeCellDivisions as the x-data.  Add the regression line to the plot using the function abline(myLm), where myLm is the linear model you created.

```{r}

```

##### c) What is the p-value for the null hypothesis that the slope of the regression between these two variables is zero?  What is the r-squared value of the model?

```{r}

```

##### d): Are the assumptions of constant variance and normal distribution of the residues reasonable for this model?  Justify your answer.

```{r}

```

***

#### 2) Consider the case-control file for the colorectal adenomas data set that is here: http://afodor.github.io/classes/stats2015/caseControlData.txt. A separate file gives obesity (BMI) data for these same subjects:http://afodor.github.io/classes/stats2015/BMI_Data.txt. For each OTU in the spreadsheet, generate a p-value from linear regression comparing BMI to the relative abundance of each OTU.    Graph out all the p-values.  Do they appear uniformly distributed? Does the microbial community appear to be influencing body weight in this cohort?  Are any of these associations significant at a 10% false discovery rate? Hints:  To lookup the ids in the BMI table, you will need to some processing on the “sample” column in the caseControl file.  The following code will convert the a sampleID so that it will match the BMI file.# remove case and control key <- sub("case", "", sampleID) key <- sub("control", "", key) # remove extraneous information from the suffix key <- strsplit( key, "_")[[1]][1]

Also, to get the p-value out of the linear model try:
	anova( myLm)$"Pr(>F)"[1]
We’ll see why that work shortly in future lectures.
