---
title: "Lab3"
author: "David Wilson • dwils152@uncc.edu"
date: "2/10/2022"
output: html_document
---

<br>

### Problem 1:

##### a) Plot the prior graph for a situation for a coin where the prior belief for p(head) is represented by the following R code : `dexp(x, rate=5) / 0.9932621`. We choose the denominator to make the Integral between 0 and 1 sum to 1.

```{r}

library(ggplot2)

theta <- seq(0.0005, 1, by=0.001)
p_theta <- dexp(theta, rate=5) / 0.9932621
density <- p_theta / sum(p_theta)

#----plot----
prior_df <- data.frame(theta, density)
ggplot(prior_df, aes(x=theta, y=density)) + geom_line(color='dodgerblue4', size=1) + xlab("Theta")+ ylab("Density") + theme(text=element_text(size=20), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"))

```

##### b) Calculate the posterior graph with both the Metropolis algorithm and grid approximation for a case with 14 heads and 10 tails (where x = prob(head)).Show the two methods roughly agree. Compare these to a plot with a posterior for new data of 14 heads and 10 tails with a prior with beta(40,40).

*Analytical Solution*
```{r}
#posterior for old prior beta 40, 40
p_theta_as <- dbeta(theta, 51, 55)
density_as <- p_theta_as/sum(p_theta_as)
analytical_df <- data.frame(theta, density_as) 

#----plot----
ggplot(analytical_df, aes(x=theta, y=density_as))+ geom_line(color='mediumpurple2', size=1) + ggtitle("Analytical Solution") + xlab("Theta") + ylab("Density") + theme(text=element_text(size=20), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"))
```

*Numerical Solution: Metropolis Algorithm*
```{r}
#prior of 9 heads and 9 tails
theta_old <- 0.5
posterior_dist_ma <- vector(length=1000000)

for (i in 1: 1000000) {
  
  p_old <- (dexp(theta_old, rate=5) / 0.9932621) * dbinom(14, 24, theta_old)
  theta_new <- theta_old + rnorm(1, 0, sd=0.02)
  
  if ( theta_new > 1)
    piNew = 1;
  
  if ( theta_new < 0 )
    piNew = 0;
  
  p_new <- (dexp(theta_new, rate=5) / 0.9932621) * dbinom(14, 24, theta_new)
  ratio <- p_new / p_old
  
  if (ratio > 1 || ratio >= runif(1) )
    theta_old = theta_new;
  
  posterior_dist_ma[i] = theta_old
  
}
#----plot----
hist <- hist(posterior_dist_ma, breaks=seq(0, 1, by=0.001), plot=FALSE)
met_alg_df <- data.frame(hist$mids, hist$counts)
ggplot(met_alg_df, aes(hist.mids, hist.counts/sum(hist.counts))) + geom_line(color='darkorange2', size=1) + ggtitle("Metropolis Algorithm") + xlab("Theta") + ylab("Density") + theme(text=element_text(size=20), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"))


```

*Numerical Solution: Grid Approximation*
```{r}

num_breaks <- 1000
posterior_dist_ga <- vector(length=num_breaks)
x_vals <- theta

i <- 1
sum <- 0

for ( x in x_vals) {
  
  posterior_dist_ga[i] <- dexp(x, rate =5) / 0.9932621 * dbinom(14, 24, x)
  sum <- sum + posterior_dist_ga[i];
  i <- i + 1;
  
}

grid_approx_df <- data.frame(x_vals, posterior_dist_ga)
ggplot(grid_approx_df, aes(x_vals, posterior_dist_ga/sum(posterior_dist_ga))) + geom_line(color='dodgerblue4', size=1) + ggtitle("Grid Approximation") + xlab("Theta") + ylab("Density") + theme(text=element_text(size=20), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"))
```

*Overlay*
```{r}

df_all <- data.frame(theta, density, hist$counts/sum(hist$counts), posterior_dist_ga)

ggplot(df_all) + geom_line(aes(x=theta, y=density_as), color='mediumpurple2', size=1) + geom_line(aes(x=theta, y=hist$counts/sum(hist$counts)), color='darkorange2', size=1) + geom_line(aes(x=theta, y=posterior_dist_ga/sum(posterior_dist_ga)), color='dodgerblue4', size=1) 
  

```

##### c)  Repeat the above calculation but for a case of 583 heads and 417 tails. (You may need toadjust your model step parameters to try and get the grid and Metropolis graphs to match up). How do the three posterior curves relate to each other now?  Why does this plot look different than the plot in (b)?

*Analytical Solution*
```{r}
#posterior for old prior beta 40, 40
p_theta_as <- dbeta(theta, 417, 583)
density_as <- p_theta_as/sum(p_theta_as)
analytical_df <- data.frame(theta, density_as) 

#----plot----
ggplot(analytical_df, aes(x=theta, y=density_as))+ geom_line(color='mediumpurple2', size=1) + ggtitle("Analytical Solution") + xlab("Theta") + ylab("Density") + theme(text=element_text(size=20), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"))
```

*Numerical Solution: Metropolis Algorithm*
```{r}

theta_old <- 0.5
posterior_dist <- vector(length=1000000)

for (i in 1: 1000000) {
  
  p_old <- (dexp(theta_old, rate=5) / 0.9932621) * dbinom(417, 583, theta_old)
  theta_new <- theta_old + rnorm(1, 0, sd=0.02)
  
  if ( theta_new > 1)
    piNew = 1;
  
  if ( theta_new < 0 )
    piNew = 0;
  
  p_new <- (dexp(theta_new, rate=5) / 0.9932621) * dbinom(417, 583, theta_new)
  ratio <- p_new / p_old
  
  if (ratio > 1 || ratio >= runif(1) )
    theta_old = theta_new;
  
  posterior_dist[i] = theta_old
  
}
#----plot----
hist <- hist(posterior_dist, breaks=seq(0, 1, by=0.001), plot=FALSE)
met_alg_df <- data.frame(hist$mids, hist$counts)
ggplot(met_alg_df, aes(hist.mids, hist.counts/sum(hist.counts))) + geom_line(color='darkorange2', size=1) + xlab("Theta") + ylab("Density") + theme(text=element_text(size=20), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"))
```

*Numerical Solution: Grid Approximation*
```{r}

num_breaks <- 1000
posterior_dist_ga <- vector(length=num_breaks)
x_vals <- theta

i <- 1
sum <- 0

for ( x in x_vals) {
  
  posterior_dist_ga[i] <- dexp(x, rate =5) / 0.9932621 * dbinom(417, 583, x)
  sum <- sum + posterior_dist_ga[i];
  i <- i + 1;
  
}

grid_approx_df <- data.frame(x_vals, posterior_dist_ga)
ggplot(grid_approx_df, aes(x_vals, posterior_dist_ga/sum(posterior_dist_ga))) + geom_line(color='dodgerblue4', size=1) + ggtitle("Grid Approximation") + xlab("Theta") + ylab("Density") + theme(text=element_text(size=20), axis.title.x=element_text(face="bold"), axis.title.y=element_text(face="bold"))
```

*Overlay*
```{r}

df_all <- data.frame(theta, density, hist$counts/sum(hist$counts), posterior_dist_ga)

ggplot(df_all) + geom_line(aes(x=theta, y=density_as), color='mediumpurple2', size=1) + geom_line(aes(x=theta, y=hist$counts/sum(hist$counts)), color='darkorange2', size=1) + geom_line(aes(x=theta, y=posterior_dist_ga/sum(posterior_dist_ga)), color='dodgerblue4', size=1) 
  

```